---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r, echo=TRUE}
library(ggplot2)
library(plyr)
library(Hmisc)
library(Rmisc)
activity <- read.table(unz("activity.zip", "activity.csv"), header=T, quote="\"", sep=",")

```


## What is mean total number of steps taken per day?
Calculating the number of steps taken per day
```{r, echo=TRUE}
attach(activity)
#tapply(steps, date, sum)
stepsperday=ddply(activity,.(activity$date),summarise,totalsteps = sum(steps,na.rm=TRUE))
tapply(steps, date, sum)

summary(stepsperday$totalsteps)
```
the mean steps are 9354

```{r, echo=TRUE}
p1=qplot(activity$steps, geom="histogram") 
p1

```



## What is the average daily activity pattern?

the code below takes the mean steps for each interval and plots the result
then sorts the data by the number of steps and lists the top 6
```{r, echo=TRUE}

timeint=ddply(activity,.(interval),summarise,mean = mean(steps,na.rm=TRUE))
plot(mean ~ interval, type = "l", data = timeint)
timeint=timeint[order(timeint$mean),]
tail (timeint)
```

## Imputing missing values

the code below first counts the number of missing values
then using ddply imputes missing values with the mean for the interval
then creates the mean steps by interval, orders the resulting dataframe with the imputed values
(timeint2) and summarizes


```{r, echo=TRUE}
sum(is.na(steps))
activity2 <- ddply(activity, "interval", mutate, imputed.value = impute(steps, mean))
timeint2=ddply(activity2,.(interval),summarise,mean = mean(steps,na.rm=TRUE))
timeint2=timeint2[order(timeint2$mean),]
tail (timeint2)

```
no change in the period with the highest steps

plot histogram with imputation (p2) compared with without imputation (p1)

```{r, echo=TRUE}

p2=qplot(activity2$steps, geom="histogram") 
stepsperday2=ddply(activity2,.(activity2$date),summarise,totalsteps2 = sum(imputed.value,na.rm=TRUE))
multiplot(p1, p2,  cols=2)

```
calculate and compare mean and median with and without imputation

```{r, echo=TRUE}
summary(stepsperday$totalsteps)
summary(stepsperday2$totalsteps2)
```


## Are there differences in activity patterns between weekdays and weekends?

convert date into days of the week
create factor indicating weekdays and weekend using chron package

```{r, echo=TRUE}

activity2$day <- weekdays(as.Date(activity2$date))
activity2$factor=as.factor(activity2$day)
library(chron)
activity2$weekend = chron::is.weekend(activity2$date)

```

finally creates dataframe with average of steps by interval separated by two levels (weekday and weekend)
plots both next to eachother

```{r, echo=TRUE}

timeint3=ddply(activity2,.(interval, weekend),summarise,mean = mean(steps,na.rm=TRUE))
timeint3$weekend=as.factor(timeint3$weekend)
timeint3$weekend=revalue(timeint3$weekend, c("TRUE"="Weekend", "FALSE"="Weekday"))

plot <- ggplot(timeint3, aes(x=interval, y=mean)) + geom_line(shape=1)

plot + facet_grid(. ~ weekend)
```
on weekdays activity is concentrated early in the mornings.
on weekends it is distributed across the day


